AFQ-Insight

### 2019-11-22 reports/manuscript/sections/introduction.tex

Using this fact, one of the most powerful methods currently available to put
macro- and micro-structure together is \emph{tractometry} (also known as
tract-based morphometry): assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are several
different available implementations of this overall idea, the principles are
similar \cite{yeatman2012, Yendiki2011-ay, Wassermann2016-iv, ODonnell2009-uu}:
tractometry begins by delineating the parts of the white matter that belong to
different major tracts and axonal fiber bundles, such as the corticospinal tract
or arcuate fasciculus, and samples the biophysical properties (such as
fractional anisotropy or mean diffusivity) along the length of these bundles. In
some cases, tissue properties along the length of each bundle are summarized by
taking the mean along each bundle, but there is a large body of evidence showing
that there is systematic variability within each bundle. This justifies
retaining the individual samples along the length of each fiber
\cite{yeatman2012, colby2012, ODonnell2009-uu}. However, while this retains
important information about each individual's white matter, it also presents
statistical challenges due to the dimensionality of the data. Based on
tractometry, researchers may choose to compare different individuals to each
other. This is usually done following one of the following approaches:

@@

In the present work, we present a novel framework for analysis of tractometry
that simultaneously selects the features for analysis, and fits a model to these
features. Rather than a comparative approach, this framework takes a predictive
approach to inference \cite{Breiman2001}. We use a linear modeling approach,
which aims to predict phenotypical variance in a group of subjects, based on a
linear combination of the features estimated with tractometry. Using this
approach, we need to first deal with the large and asymmetric dimensionality of
the data: tractometry data usually has many more features (i.e., number of
measurements per individual) than samples (number of subjects), which makes
inferences from the data about phenotypical differences between individuals
ill-posed. This regime is the target of several statistical learning techniques,
and is often solved by various forms of regularization. For example, Tikhonov
regularization shrinks the solution such that the sum of squared contributions
from the individual features are minimized \cite{Hoerl2000-ij}. Another solution
to the problem is provided by the Lasso algorithm, which instead minimizes the
sum of the absolute values of contributions of each feature
\cite{Tibshirani1996-qs}. This tends to shrink to zero the contributions of many
of the features, providing results that are both accurate and interpretable.
When additional structure is available in the organization of the data,
regularization algorithms can take advantage of this structure. For example, if
the features lend themselves to a natural division into different groups, the
group lasso (GL) can be used to select groups of features, rather than
individual features \cite{Yuan2006-ky}. The Sparse Group Lasso (SGL) elaborates
on this idea by providing control both of group sparsity, as well as overall
sparsity of the solutions \cite{simon2013sgl}. Because the features measured
with tractomery lend themselves to grouping based on the tracts from which each
measurement is taken, GL and SGL provide a useful tool linear model fitting in
problems of this form. Using linear regression as the basis for analysis is
useful for both regression tasks, in which a continuous measure (e.g., age) is
modeled, as well as for classification tasks, where a discrete property is
modeled (e.g., ). Moreover, use of SGL in particular helps derive informative
properties of the human white matter, both in cases where the underlying causes
of variance are sparsely concentrated (e.g., in one particular white matter
tract) and in cases where biological substrates of differences between
individuals are more widespread. This is because SGL is controlled by
meta-parameters that set the degree to  which and that can be set empirically,
based on the data, in a process of nested cross-validation. We demonstrate these
properties of the algorithm in data from previously published studies
\cite{sarica2017corticospinal, yeatman2014lifespan}.

>>>>

Using this fact, one of the most powerful methods currently available to put
macro- and micro-structure together is \emph{tractometry} (also known as
tract-based morphometry): assessment of the physical properties of the white
matter along specific tracts \cite{Bells2011-cf}. Though there are several
different available implementations of this overall idea, the principles are
similar \cite{yeatman2012tract, Yendiki2011-ay, Wassermann2016-iv, ODonnell2009-uu}:
tractometry begins by delineating the parts of the white matter that belong to
different major tracts and axonal fiber bundles, such as the corticospinal tract
or arcuate fasciculus, and samples the biophysical properties (such as
fractional anisotropy or mean diffusivity) along the length of these bundles. In
some cases, tissue properties along the length of each bundle are summarized by
taking the mean along each bundle, but there is a large body of evidence showing
that there is systematic variability within each bundle. This justifies
retaining the individual samples along the length of each fiber
\cite{yeatman2012tract, colby2012, ODonnell2009-uu}. However, while this retains
important information about each individual's white matter, it also presents
statistical challenges due to the dimensionality of the data. Based on
tractometry, researchers may choose to compare different individuals to each
other. This is usually done following one of the following approaches:

@@

In the present work, we present a novel framework for analysis of tractometry
that simultaneously selects the features for analysis, and fits a model to these
features. Rather than a comparative approach, this framework takes a predictive
approach to inference \cite{Breiman2001-uz}. We use a linear modeling approach,
which aims to predict phenotypical variance in a group of subjects, based on a
linear combination of the features estimated with tractometry. Using this
approach, we need to first deal with the large and asymmetric dimensionality of
the data: tractometry data usually has many more features (i.e., number of
measurements per individual) than samples (number of subjects), which makes
inferences from the data about phenotypical differences between individuals
ill-posed. This regime is the target of several statistical learning techniques,
and is often solved by various forms of regularization. For example, Tikhonov
regularization shrinks the solution such that the sum of squared contributions
from the individual features are minimized \cite{Hoerl2000-ij}. Another solution
to the problem is provided by the Lasso algorithm, which instead minimizes the
sum of the absolute values of contributions of each feature
\cite{Tibshirani1996-qs}. This tends to shrink to zero the contributions of many
of the features, providing results that are both accurate and interpretable.
When additional structure is available in the organization of the data,
regularization algorithms can take advantage of this structure. For example, if
the features lend themselves to a natural division into different groups, the
group lasso (GL) can be used to select groups of features, rather than
individual features \cite{Yuan2006-ky}. The Sparse Group Lasso (SGL) elaborates
on this idea by providing control both of group sparsity, as well as overall
sparsity of the solutions \cite{simon2013sgl}. Because the features measured
with tractomery lend themselves to grouping based on the tracts from which each
measurement is taken, GL and SGL provide a useful tool linear model fitting in
problems of this form. Using linear regression as the basis for analysis is
useful for both regression tasks, in which a continuous measure (e.g., age) is
modeled, as well as for classification tasks, where a discrete property is
modeled (e.g., ). Moreover, use of SGL in particular helps derive informative
properties of the human white matter, both in cases where the underlying causes
of variance are sparsely concentrated (e.g., in one particular white matter
tract) and in cases where biological substrates of differences between
individuals are more widespread. This is because SGL is controlled by
meta-parameters that set the degree to  which and that can be set empirically,
based on the data, in a process of nested cross-validation. We demonstrate these
properties of the algorithm in data from previously published studies
\cite{sarica2017corticospinal, yeatman2014lifespan}.
